---
title: "Dorsey Assignment 2"
output: html_document
date: "2025-10-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


pacman::p_load(dplyr, caTools, tm, mlapi, e1071, caret, here, stringr, class)

```


Read in the data sources 
```{r}
train <- read.csv(here("train.csv"))
test <- read.csv(here("test_without_labels.csv"))
dev<- read.csv(here("dev.csv"))
```




##------Aspect-Orientation--------##


To customize the classification strategy to address aspect-oriented sentiment classification, take the drug mentioned in the therapy column and replace its mention in the Text column with the generic keyword _THERAPY_. Use word boundaries to make sure only the full word is flagged for replacement - no variations 


```{r}

#Training data set 

train <- train %>%
  rowwise () %>% 
  mutate(text = str_replace_all(text, 
                                regex(paste0("\\b", therapy,"\\b"), ignore_case=TRUE), 
                                "_THERAPY_")) %>%
  ungroup()

#Dev data set 

dev <- dev %>%
  rowwise () %>% 
  mutate(text = str_replace_all(text, 
                                regex(paste0("\\b", therapy, "\\b"), ignore_case=TRUE), 
                                "_THERAPY_")) %>%
  ungroup()

```


##--------Preprocess the data-------##

Create a corps, make everything lowercase, remove punctuation, remove common English stopwords, stem words
```{r}
#Training data
train_text <-train$text
train_text_corpus <- VCorpus(VectorSource(train_text))
train_text_corpus <- tm_map(train_text_corpus, content_transformer(tolower))
train_text_corpus <- tm_map(train_text_corpus, removePunctuation)
train_text_corpus <- tm_map(train_text_corpus, removeWords,stopwords("english"))
train_text_corpus <- tm_map(train_text_corpus, stemDocument)
length(train_text_corpus)


#Dev data
dev_text <-dev$text
dev_text_corpus <- VCorpus(VectorSource(dev_text))
dev_text_corpus <- tm_map(dev_text_corpus, content_transformer(tolower))
dev_text_corpus <- tm_map(dev_text_corpus, removePunctuation)
dev_text_corpus <- tm_map(dev_text_corpus, removeWords,stopwords("english"))
dev_text_corpus <- tm_map(dev_text_corpus, stemDocument)
length(dev_text_corpus)
```
Generate n-grams in each corpus (unigrams, bigrams, trigrams)

```{r}
#Tokenizer function 
NLP_tokenizer <- function(x) {
  unlist(lapply(ngrams(words(x), 1:3), paste, collapse = "_"), use.names = FALSE)
}

#Tokenize Training data 
n_gram_train_corpus <- tm_map(train_text_corpus,content_transformer(NLP_tokenizer))

#Tokenize Dev data 
n_gram_dev_corpus <- tm_map(dev_text_corpus,content_transformer(NLP_tokenizer))

#Verify this worked: 
n_gram_train_corpus[[3]]$content
```
###---------Split the data------------###

The Training and evaluation split is already performed

```{r}
#Extract the sentiment labels in the training and eval set
train_classes <- train$label
dev_classes <- dev$label


#However, we still want to split the training set for hyperparameter optimization. Use an 80/20 split
set.seed(123)
split<- sample.split(train_classes, SplitRatio = 0.8)
n_gram_train_corpus_split<- subset(n_gram_train_corpus, split == TRUE)
train_classes_split <- subset(train_classes, split == TRUE)
validation_ngram_corpus <- subset(n_gram_train_corpus, split == FALSE) 
validation_classes <- subset(train_classes, split == FALSE)
```


###-------Vectorize the training data (60%)-----------###

Creating a document term matrix and removing sparse terms

```{r}
training_dct_matrix <- DocumentTermMatrix(n_gram_train_corpus_split)
training_dct_matrix_ <- removeSparseTerms(training_dct_matrix, 0.995)

inspect(training_dct_matrix_)

```

###-------Vectorize the validation data-----------###

Use the dictionary from the training document matrix!

```{r}
val_dct_matrix_ <- DocumentTermMatrix(validation_ngram_corpus, list(dictionary=colnames(training_dct_matrix_)))
```



Convert to R data frames 
```{r}
training_matrix_df <- as.data.frame(as.matrix(training_dct_matrix_))
val_matrix_df<- as.data.frame(as.matrix(val_dct_matrix_))
colnames(training_matrix_df) <- make.names(colnames(training_matrix_df))
colnames(val_matrix_df) <- make.names(colnames(val_matrix_df))
training_matrix_df$label <- train_classes_split
training_matrix_df$label <- as.factor(training_matrix_df$label)

```



###---------Optimize Hyperparameters--------###

For SVm, I want to tune the following:  
  -cost (1,2,4,8,16, 32, 64, 128, 256)
  -polynomial v. radial basis function 
  
For Random Forest, I want to tune the following: 
  -Number of trees (10, 50, 100, 500, 1000, 2500, 5000, 7500, 10000)
  
  
I want to calculate the overall accuracy and calculate the F1 score among the positive and negative classes for all hyperparameter options

```{r}
#SVM where kernel = polynomial 

i = 1
while (i<= 256) {
  cat("==============\n")  
cat("cost =", i, "\n")
cat("==============\n")
  trained_model <- svm(label~., data=training_matrix_df, cost=i, kernel = "polynomial")
  predictions<- predict(trained_model, newdata = val_matrix_df)
  print(i)
cm <-  confusionMatrix(as.factor(validation_classes), predictions)

print(cm)

sens <-cm$byClass[,"Sensitivity"]
ppv <- cm$byClass[, "Pos Pred Value"]

f1 <- (2 * sens * ppv)/(sens+ppv) 

cat("\nF1 Scores by Class:\n")
print(f1)

i = i*2
  
}

#SVM where kernel = radial

i = 1
while (i<=256) {

cat("==============\n")  
cat("cost =", i, "\n")
cat("==============\n")

trained_model <- svm(label~., data=training_matrix_df, cost=i, kernel = "radial")
predictions<- predict(trained_model, newdata = val_matrix_df)
print(i)
cm <- confusionMatrix(as.factor(validation_classes), predictions)
  
print(cm)

sens <-cm$byClass[,"Sensitivity"]
ppv <- cm$byClass[, "Pos Pred Value"]

f1 <- (2 * sens * ppv)/(sens+ppv) 

cat("\nF1 Scores by Class:\n")
print(f1)

i = i*2
}


#Random forest with different numbers of trees 

library(randomForest)

tree_val <- c(10, 50, 100, 500, 1000, 2500, 5000, 7500, 10000)

rf_results<- list()

for (n in tree_val){
  
cat("==============\n")  
cat("ntree =", n, "\n")
cat("==============\n")


trained_model_rf <- randomForest(label ~ ., ntree = n, data = training_matrix_df)

predictions_rf <- predict(trained_model_rf, newdata=val_matrix_df)

cm <-confusionMatrix(as.factor(validation_classes), predictions_rf)

 
print(cm)

sens <-cm$byClass[,"Sensitivity"]
ppv <- cm$byClass[, "Pos Pred Value"]

f1 <- (2 * sens * ppv)/(sens+ppv) 

cat("\nF1 Scores by Class:\n")
print(f1)


}



```

###--------Select the Optimized Model----------###

Evaluation suggests that the optimal model is an SVM model with kernel = radial and cost = 8




###-----------Evaluate performance on Dev dataset------------###

```{r}
#vectorize the test set. Use the column names from the training dictionary
dev_dct_matrix_ <- DocumentTermMatrix(n_gram_dev_corpus, list(dictionary=colnames(training_dct_matrix_)))

#Convert to data frames 
dev_matrix_df <- as.data.frame(as.matrix(dev_dct_matrix_))
colnames(dev_matrix_df) <- make.names(colnames(dev_matrix_df))

```


Train the model and generate predictions
```{r}
trained_model_full <- svm(label~., data=training_matrix_df, cost = 8, kernel = "radial")
predictions_full <- predict(trained_model_full, newdata = dev_matrix_df)
cm_final <-confusionMatrix(as.factor(dev_classes), predictions_full)
print(cm_final)


sens_final <-cm_final$byClass[,"Sensitivity"]
ppv_final <- cm_final$byClass[, "Pos Pred Value"]

f1_final <- (2 * sens_final * ppv_final)/(sens_final+ppv_final) 
f1_final

```

###---------Classify Unlabeled Data----------###

Need to use the Test dataset 

Preproccess the test data 


```{r}

#Make the training aspect oriented
test <- test %>%
  rowwise () %>% 
  mutate(text = str_replace_all(text, 
                                regex(paste0("\\b", therapy,"\\b"), ignore_case=TRUE), 
                                "_THERAPY_")) %>%
  ungroup()



#Preprocess the training data
test_text <-test$text
test_text_corpus <- VCorpus(VectorSource(test_text))
test_text_corpus <- tm_map(test_text_corpus, content_transformer(tolower))
test_text_corpus <- tm_map(test_text_corpus, removePunctuation)
test_text_corpus <- tm_map(test_text_corpus, removeWords,stopwords("english"))
test_text_corpus <- tm_map(test_text_corpus, stemDocument)
length(test_text_corpus)

#Generate n grams 

n_gram_test_corpus <- tm_map(test_text_corpus,content_transformer(NLP_tokenizer))
n_gram_test_corpus[[3]]$content
```

Vectorize the test set. Use the dictionary from the training document matrix

```{r}
test_dct_matrix_ <- DocumentTermMatrix(n_gram_test_corpus, list(dictionary=colnames(training_dct_matrix_)))
```



Convert to R data frames 
```{r}
test_matrix_df<- as.data.frame(as.matrix(test_dct_matrix_))
colnames(test_matrix_df) <- make.names(colnames(test_matrix_df))
```



```{r}
predictions_new <- predict(trained_model_full, newdata = test_matrix_df)
```

Build the predictions back into the dataset

```{r}
test$label <- predictions_new

```


Identify the distribution of therapies among the newly classified test data

```{r}
sort(table(test$therapy), decreasing = TRUE)

#Most common: adderall, alprazolam, cannabis, oxycodone, oxycontin 
```

Explore the sentiment distributions associated with each of the most popular therapies 

```{r}

#Adderall: mostly neutral
table(test$label[test$therapy == "adderall"])
prop.table(table(test$label[test$therapy == "adderall"]))

#alprazolam: mostly neutral
table(test$label[test$therapy == "alprazolam"])
prop.table(table(test$label[test$therapy == "alprazolam"]))

#cannabis: mostly neutral
table(test$label[test$therapy == "cannabis"])
prop.table(table(test$label[test$therapy == "cannabis"]))



#Visualize these distributions 

ggplot(subset(test, therapy %in% c("adderall", "alprazolam", "cannabis")), aes(x = therapy, fill = label))+
  geom_bar(position = "fill")+
  scale_fill_manual(
    values = c("negative" = "red", "neutral" = "grey", "positive" = "forestgreen")) +
  theme_minimal() +
  labs(x = "Therapy", y = "Proportion of Tweets", fill = "Sentiment")



```



